Draft Manuscript

### 2022-09-23 Prereg_and_manuscript/manuscript.Rmd
Added code to methods and finished method section

**Data download**

@@

**Data preparation**

@@

@@

@@

**Plotting**

I constructed box plots using using ggplot2::ggplot

@@

>>>>

**Data download (script: "part-1_data_download")**

@@

```{r include=TRUE}
# creating directory to store data
dir.create("data")

## downloading from dryad url ----

data.url <- "https://datadryad.org/stash/downloads/file_stream/41575"
metadata.url <- "https://datadryad.org/stash/downloads/file_stream/41576"

data.dest.file <- paste0("data/","Data_ThriffaultEtAl_EcoEvo_",Sys.Date(),".zip")
metadata.dest.file <- paste0("data/","Metadata_ThriffaultEtAl_EcoEvo_",Sys.Date(),".txt")

unzip(download.file(url = data.url, destfile = data.dest.file),exdir = here::here("data"))
download.file(url= metadata.url, destfile = metadata.dest.file)

unzip(list.files(path = here::here("data"), pattern = "*.zip", full.names = TRUE),exdir = here::here("data"))

**Data preparation (script: "part-2_data_preparation")**

@@

```{r}
# Loading data tables ----
X_mat <- read.csv(here::here("data","Matrice_X.csv"),
                  sep = ";") # explanatory variables - see metadata
Y_mat <- read.csv(here::here("data","Matrice_Y.csv"),
                  sep = ";") # response variables (mean ericaceous cover, in %)
clim <- read.csv(here::here("data","Climate_value.csv"),
                 sep = ";") # numerical codes used for clim vars in X_mat

# Classifying Forest Type ----
# Identifying two most dominant tree species in each plot and assigning forest type as 
  # conifer (c)
  # deciduous (d)
  # mixed (m)
# looping through each row of X_mat dataframe

for (i in 1:dim(X_mat)[1]){
  most_abun <-  as.numeric(Rfast::nth(as.matrix(X_mat[i,2:14]), k=1, descending = T,index.return = T)) # returning the index of the column with the most abundant tree species for each row
  sec_abun <- as.numeric(Rfast::nth(as.matrix(X_mat[i,2:14]), k=2, descending = T,index.return = T)) # returning index of second most abundant species
  dominant_sp <- c(most_abun, sec_abun)+1 # adding 1 to account for the first column which was left out of vector for previous function
  dominant_sp <- colnames(X_mat)[dominant_sp] # creating character vector containing column names of most dominant species

  if (dominant_sp[1] == "AB" || # if both species conifers
      dominant_sp[1] == "PB" ||
      dominant_sp[1] == "PM"||
      dominant_sp[1] == "PM_AB" ||
      dominant_sp[1] == "TO" &&
      dominant_sp[2] == "AB" ||
      dominant_sp[2] == "PB" ||
      dominant_sp[2] == "PM"||
      dominant_sp[2] == "PM_AM" ||
      dominant_sp[2] == "TO"){
      X_mat$Forest_Type[i] = "c" # forest type conifer

  }else if (dominant_sp[1] == "BA" || # if both species deciduous
            dominant_sp[1] == "BP" ||
            dominant_sp[1] == "PT" &&
            dominant_sp[2] == "BA" ||
            dominant_sp[2] == "BP" ||
            dominant_sp[2] == "PT"){
    X_mat$Forest_Type[i] = "d" # forest type deciduous

  }else if (dominant_sp[1] == "BP_AB" || # if either species is mixed
            dominant_sp[1] == "BP_PM" ||
            dominant_sp[1] == "PT_AB"|| 
            dominant_sp[1] == "PT_PM" ||
            dominant_sp[1] == "BP_AB" ||
            dominant_sp[1] == "BP_PM" ||
            dominant_sp[1] == "PT_AB"|| 
            dominant_sp[1] == "PT_PM"){
    X_mat$Forest_Type[i] = "m"  # forest type mixed

  }else if(dominant_sp[1] == "AB" || # if first species is conifer, second is deciduous 
            dominant_sp[1] == "PB" ||
            dominant_sp[1] == "PM"||
            dominant_sp[1] == "PM_AB" ||
            dominant_sp[1] == "TO" &&
            dominant_sp[2] == "BA" ||
            dominant_sp[2] == "BP" ||
            dominant_sp[2] == "PT"){
    X_mat$Forest_Type[i] = "m" # forest type mixed

  }else if(dominant_sp[1] == "BA" || # if first species is deciduous, second is conifer
            dominant_sp[1] == "BP" ||
            dominant_sp[1] == "PT" &&
            dominant_sp[2] == "AB" ||
            dominant_sp[2] == "PB" ||
            dominant_sp[2] == "PM"||
            dominant_sp[2] == "PM_AB" ||
            dominant_sp[2] == "TO"){
            X_mat$Forest_Type[i] = "m" # forest type mixed
}else{
  X_mat$Forest_Type[i] = "n" # non-forested - other
}
}

# checking the dimensions of the dataframe when we select for specific cateogires
# tells us that the conifer type dominates
# this difference will be normalized later when we take the difference

dim(dplyr::filter(X_mat,Forest_Type=="c")) 
dim(dplyr::filter(X_mat,Forest_Type=="d")) 
dim(dplyr::filter(X_mat,Forest_Type=="m"))
dim(dplyr::filter(X_mat,Forest_Type=="n"))

@@

```{r}
# Classifying as high or low precipitation ----
# PRECU is the annual rainfall variable

med <- median(X_mat$PRECU)
# will use the median as the threshold

X_mat <- X_mat %>% 
  mutate(., RainFall = case_when(PRECU < med ~ "L", 
            PRECU >= med ~ "H"))

@@

```{r}
# Combining and classifying as forest type - rainfall index
X_mat <- 
  X_mat %>% 
  unite(., rf_index, c("RainFall","Forest_Type"), sep = "_", remove =F, na.rm = FALSE)

# writing altered data as .csv ----
write.csv(X_mat, here::here("data","X_mat_processed.csv"))
```

**Plotting (script: "part-3_data_analysis")**

I constructed box plots displaying percent cover of each species as the response variable and rf (rainfall forest cover index) as the grouping variable using using ggplot2::ggplot. I decided to log-transform the response variable since the data was skewed with outliers making the display and comparison between groups difficult in all of the plots.

```{r}
# loading in prepared data ----
  X_mat <- read.csv(here::here("data","X_mat_processed.csv"))
  X_mat <- X_mat %>% 
    select(CLE, Forest_Type, PRECU, RainFall,rf_index)

  Y_mat <- read.csv(here::here("data","Matrice_Y.csv"),
                    sep = ";") # response variables (mean ericaceous cover, in %)
  clim <- read.csv(here::here("data","Climate_value.csv"),
                   sep = ";") # numerical codes used for clim vars in X_mat

# PLOTTING WITH BOX PLOTS----

  # For Chamaedaphne calyculata
  CAL <- data.frame(Y_mat$V_CAL,X_mat$PRECU, X_mat$rf_index) # creating dataframe with only this species
  colnames(CAL) <- c("V_CAL","PRECU","rf_index") # renaming columns 
  ggplot(data=CAL, aes(x=rf_index, y=V_CAL)) + # plotting only this species with the rainfall-forest type index in box plots
    geom_boxplot()

  ggplot(data=CAL, aes(x=rf_index, y=log(V_CAL))) + # log transforming the y axis for easier pattern detection
    geom_boxplot()  # automatically removes observations with 0 % cover - which is ok

  # For Rhododendron groenlandicum
  RHG <- data.frame(Y_mat$V_RHG,X_mat$PRECU,X_mat$rf_index) # creating dataframe with only this species
  colnames(RHG) <- c("V_RHG","PRECU","rf_index") # renaming columns 
  ggplot(data=RHG, aes(x=rf_index, y=V_RHG)) + # plotting only this species with the rainfall-forest type index in box plots
    geom_boxplot()

  ggplot(data=RHG, aes(x=rf_index, y=log(V_RHG))) + # log transforming the y axis for easier pattern detection
    geom_boxplot()  # automatically removes observations with 0 % cover - which is ok

## For Kalmia angustifolia
  # Plotting: 
  KAA <- data.frame(Y_mat$V_KAA,X_mat$PRECU,X_mat$rf_index) # creating dataframe with only this species
  colnames(KAA) <- c("V_KAA","PRECU","rf_index") # renaming columns

  ggplot(data=KAA, aes(x=rf_index, y=V_KAA)) + # plotting only this species with the rainfall-forest type index in box plots
    geom_boxplot()

  ggplot(data=KAA, aes(x=rf_index, y=log(V_KAA))) + # log transforming the y axis for easier pattern detection
    geom_boxplot()  # automatically removes observations with 0 % cover - which is ok

## For Vaccinium sp. 
    VAAM <- data.frame(Y_mat$V_VAAM,X_mat$PRECU,X_mat$rf_index) # creating dataframe with only this species
    colnames(VAAM) <- c("V_VAAM","PRECU","rf_index") # renaming columns

    ggplot(data=VAAM, aes(x=rf_index, y=V_VAAM)) + # plotting only this species with the rainfall-forest type index in box plots
      geom_boxplot()

    ggplot(data=VAAM, aes(x=rf_index, y=log(V_VAAM))) + # log transforming the y axis for easier pattern detection
      geom_boxplot()  # automatically removes observations with 0 % cover - which is ok

I then combined all of the above plots into one single chart for easier comparison.

```{r}
## Plotting altogether in facets 
    # creating new variable to denote species to make table with all species
    CAL <- CAL %>% 
      add_column("species" = "CAL") %>%  # adding column with species names
      rename(cover ="V_CAL") # renaming percent cover column to represent all species

    RHG <- RHG %>% 
      add_column("species" = "RHG") %>% 
      rename(cover ="V_RHG")

    KAA <- KAA %>% 
      add_column("species" = "KAA") %>% 
      rename(cover ="V_KAA")

    VAAM <- VAAM %>% 
      add_column("species" = "VAAM") %>% 
      rename(cover ="V_VAAM")

    plot_cover <- rbind(CAL,RHG, KAA, VAAM) # combining all species by rows into one table

    ## creating ggplot object
    p <- ggplot(data= plot_cover, aes(x=rf_index, y=log(cover))) + # log transforming the y axis for easier pattern detection
      geom_boxplot()+
      xlab("Rainfall-Forest Type Index (RF)") + # changing x axis labe
      ylab("Log (% cover)")+ # changing y axis label
      theme(axis.text.x = element_text(angle = 45, vjust = 0.5), # changing plot themes for angled text and no gridlines
            panel.grid = element_blank())

    ## function to changes names to each plot in facet 
    # from https://stackoverflow.com/questions/3472980/how-to-change-facet-labels
      species_names <- list(
        'CAL'="C. calyculata",
        'KAA'="K. angustifolia",
        'RHG'="R. groenlandicum",
        'VAAM'="Vaccinium sp."
      )

      species_labeller <- function(variable,value){
        return(species_names[value])
      }

    ## creating facet plot with all species plots in one graph 
    p + facet_wrap( ~ species, scales="free",
                    labeller=species_labeller)

    # exporting plot to jpeg
    jpeg(file=here::here("outputs","Cover-RF_Box-Plot.jpeg"))
    p + facet_wrap( ~ species, scales="free",
                    labeller=species_labeller)
    dev.off()
```

**Statistical analysis**

I used a Kruksal-Wallis test to determine if any any groups were significantly different from the rest, or in other words, if at least one rainfall forest cateogry contains a significantly different percent cover of a given species than the other groups. For all species, p \<0.05 indicating we can reject the null hypothesis that percent cover is the same across rainfall forest type categories for all species. To further assess the relationships, I performed a pairwise wilcox test to determine whether percent cover differed significantly from each other in pairs. This was not in the initial pre-registration but is an extended step to further test the hypothesis.

```{r}
 # We want to know is there any significant difference between percent over of plants in different rainfall and forest type conditions

    ## 1) C. calyculata
      # Kruksal-Wallis test to determine if significant difference between means of rf-index groups
      CAL_cover <- plot_cover %>% filter(species=="CAL") # creating daframe with only this species
      kruskal.test(cover ~ rf_index, data = CAL_cover) # performing test - P < 0.05 - yes, indicates significatn differences between groups 
      # use pairwise wilcox test to determine which groups different 
      CAL_wt <- pairwise.wilcox.test(CAL_cover$cover, CAL_cover$rf_index,
                           p.adjust.method = "BH")
      CAL_wt_pframe <- data.frame(CAL_wt[["p.value"]]) # taking p-value matrix and coercing to dataframe
      write.csv(CAL_wt_pframe, here::here("outputs","Ccalyculata_pframe.csv")) # saving dataframe as csv
      write.csv(CAL_wt_pframe>=0.05, here::here("outputs","Ccalyculata_pframe_binary.csv")) # saving data frame indicating p< 0.05 for each paru

      # Creating matrix of difference between medians to pair with the p-value table 
      CAL_meds <- CAL_cover %>% # summarizing by rf group
        group_by(rf_index) %>% 
        summarise(
          median = median(cover, na.rm = TRUE) # calculating median for each group
        )

      CAL_meds_diff <- as.data.frame(outer(CAL_meds$median, CAL_meds$median, FUN= "-")) # creating pairwise difference matrix between medians
      colnames(CAL_meds_diff) <- CAL_meds$rf_index # renaming to have rf index names
      rownames(CAL_meds_diff) <- CAL_meds$rf_index
      write.csv(CAL_meds_diff, here::here("outputs","Ccalyculata_med_differences.csv")) # saving this file as .csv

    ##2) Repeating for K. angustifolia
      # Kruksal-Wallis test to determine if significant difference between means of rf-index groups
      KAA_cover <- plot_cover %>% filter(species=="KAA") # creating daframe with only this species
      kruskal.test(cover ~ rf_index, data = KAA_cover) # performing test - P < 0.05 - yes, indicates significatn differences between groups 
      # use pairwise wilcox test to determine which groups different 
      KAA_wt <- pairwise.wilcox.test(KAA_cover$cover, KAA_cover$rf_index,
                           p.adjust.method = "BH")
      KAA_wt_pframe <- data.frame(KAA_wt[["p.value"]])
      write.csv(KAA_wt_pframe, here::here("outputs","Kangust_pframe.csv"))
      write.csv(KAA_wt_pframe>=0.05, here::here("outputs","Kangust_pframe_binary.csv"))

      # Creating matrix of difference between medians to pair with the p-value table 
      KAA_meds <- KAA_cover %>% 
        group_by(rf_index) %>% 
        summarise(
          median = median(cover, na.rm = TRUE)
        )

      KAA_meds_diff <- as.data.frame(outer(KAA_meds$median, KAA_meds$median, FUN= "-"))
      colnames(KAA_meds_diff) <- KAA_meds$rf_index
      rownames(KAA_meds_diff) <- KAA_meds$rf_index
      write.csv(KAA_meds_diff, here::here("outputs","Kangust_med_differences.csv"))

    ##3) R. groenlandicum
      # Kruksal-Wallis test to determine if significant difference between means of rf-index groups
      RHG_cover <- plot_cover %>% filter(species=="RHG") # creating daframe with only this species
      kruskal.test(cover ~ rf_index, data = RHG_cover) # performing test - P < 0.05 - yes, indicates significatn differences between groups 
      # use pairwise wilcox test to determine which groups different 
      RHG_wt <- pairwise.wilcox.test(RHG_cover$cover, RHG_cover$rf_index,
                           p.adjust.method = "BH")
      RHG_wt_pframe <- data.frame(RHG_wt[["p.value"]])
      write.csv(RHG_wt_pframe, here::here("outputs","Rgroenlandicum_pframe.csv"))
      write.csv(RHG_wt_pframe>=0.05, here::here("outputs","Rgroenlandicum_pframe_binary.csv"))

      # Creating matrix of difference between medians to pair with the p-value table 
      RHG_meds <- RHG_cover %>% 
        group_by(rf_index) %>% 
        summarise(
          median = median(cover, na.rm = TRUE)
        )

      RHG_meds_diff <- as.data.frame(outer(RHG_meds$median, RHG_meds$median, FUN= "-"))
      colnames(RHG_meds_diff) <- RHG_meds$rf_index
      rownames(RHG_meds_diff) <- RHG_meds$rf_index
      write.csv(RHG_meds_diff, here::here("outputs","Rgroenlandicum_med_differences.csv"))

    ##4) Vaccinium sp.
      # Kruksal-Wallis test to determine if significant difference between means of rf-index groups
      VAAM_cover <- plot_cover %>% filter(species=="VAAM") # creating daframe with only this species
      kruskal.test(cover ~ rf_index, data = VAAM_cover) # performing test - P < 0.05 - yes, indicates significatn differences between groups 
      # use pairwise wilcox test to determine which groups different 
      VAAM_wt <- pairwise.wilcox.test(VAAM_cover$cover, VAAM_cover$rf_index,
                           p.adjust.method = "BH")
      VAAM_wt_pframe <- data.frame(VAAM_wt[["p.value"]])
      write.csv(VAAM_wt_pframe, here::here("outputs","Vaccinium_pframe.csv"))
      write.csv(VAAM_wt_pframe>=0.05, here::here("outputs","Vaccinium_pframe_binary.csv"))

      # Creating matrix of difference between medians to pair with the p-value table 
      VAAM_meds <- VAAM_cover %>% 
        group_by(rf_index) %>% 
        summarise(
          median = median(cover, na.rm = TRUE)
        )

      VAAM_meds_diff <- as.data.frame(outer(VAAM_meds$median, VAAM_meds$median, FUN= "-"))
      colnames(VAAM_meds_diff) <- VAAM_meds$rf_index
      rownames(VAAM_meds_diff) <- VAAM_meds$rf_index

@@

\# CAL significant differences (p\< 0.05): H_d vs H_c ; H_m vs H_d ; L_c vs H_d ; L_m vs H_c ; L_m vs L_c

\# KAA significant differences (P\<0.05): H_c vs H_d ; H_c vs H_m ; H_c vs L_c ; H_d vs H_n ; H_m vs H_n ; H_m vs L_m ; H_n vs L_c ; H_n vs L_m

\# RHGsignificant differences (P\<0.05): all types with H_c; H_d vs H_m ; H_d vs H_n ; H_d vs L_c ; L_c vs L_d ; L_c vs L_m ; L_c vs L_n

\# VAAM significant differences (P\<0.05): H_c vs H_d; H_c vs H_m; H_c vs L_c ; H_c vs L_m; H_d vs L_c; H_d vs L_n ; H_m vs L_c; H_m vs L_n
